---
title: "04_sample_by_sample_analysis"
output: html_document
date: "2025-04-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries

```{r echo=FALSE}
library(dplyr)
library(Seurat)
library(patchwork)
library(here)
library(ggplot2)
library(stringr)
```

## Analysis

19/04/2025 - In this analysis, I am looking at the samples separately

```{r}
h5_dir <- here("/Users/tharveshmoideliyakatali/Documents/XX_Steffan_UiO_RNA_seq/mouse_single_cell/individual_files")
h5_files <- list.files(h5_dir, pattern = "\\.h5$", full.names = TRUE)

#Read metadata
meta_path <- here("../aggregation.csv")
metadata <- readr::read_csv(meta_path)
metadata$molecule_h5 <- NULL
```


## Reading count files

```{r}
seurat_list <- list()

for(file in h5_files)
{
  sample_name <- gsub(".h5$", "", basename(file))
  sample_name <- gsub("^[^_]+_", "", sample_name)
  counts <- Read10X_h5(file)
  seurat_obj <- CreateSeuratObject(counts = counts, project = sample_name)
  seurat_list[[sample_name]] <- seurat_obj
}
```
## Add metadata to the metadata

```{r}
for (sample_name in names(seurat_list)) {
  # Get corresponding metadata for this sample
  # Adjust this based on how your metadata is organized
  sample_metadata <- metadata %>% filter(sample_name == sample_id)
  
  # Add sample-level metadata to Seurat object
  for (col in colnames(sample_metadata)) {
    if (col != "sample") {
      seurat_list[[sample_name]][[col]] <- sample_metadata[[col]][1]
    }
  }
}
```

```{r}
for (sample_name in names(seurat_list)) {
  cat(sample_name, " ",ncol(seurat_list[[sample_name]]), "\n")
}
```

# Cell/barcode counts at the beginning

```{r}
cb_counts_start <- data.frame(
  sample = names(seurat_list),
  n_cb = sapply(seurat_list, function(obj) ncol(obj))
)
cb_counts_start
```


```{r}
min_features <- 200
min_counts <- 500

seurat_list_feature_count_filtered <- lapply(seurat_list, function(obj) {
  obj <- subset(obj, subset = nFeature_RNA > min_features & nCount_RNA > min_counts)
})
```

# Cell/barcode counts after minimum count filtering

```{r}
cb_counts_count_filtered <- data.frame(
  sample = names(seurat_list_feature_count_filtered),
  n_cb = sapply(seurat_list_feature_count_filtered, function(obj) ncol(obj))
)
cb_counts_count_filtered
```

# Join the number of cells table

```{r}
cell_count_number <- left_join(cb_counts_start, cb_counts_count_filtered, by = 'sample')
colnames(cell_count_number) <- c("sample", "cb_wo_filter", "cb_filter_by_nfeatures_ncount")
write.table(cell_count_number, file = "01_mito_content_analysis/cell_feature_umi_filtered.tsv", quote = FALSE,
            sep = "\t", row.names = FALSE)
```

```{r}
cell_count_number$cb_filter_by_nfeatures_ncount %>% mean()
```

# Calculate percent mt on the count filtered seurat object

```{r}
for (sample_name in names(seurat_list_feature_count_filtered)) {
  seurat_list_feature_count_filtered[[sample_name]][["percent.mt"]] <- PercentageFeatureSet(seurat_list_feature_count_filtered[[sample_name]], pattern = "^mt-")
}
```

# Since it is a Flex kit, I want to look at the number of mitochondiral genes in the feature list

```{r}
# Check how many features (genes) are mitochondrial
mt_genes <- grep("^mt-", rownames(seurat_obj), value = TRUE)  # Or "^MT-" if uppercase

length(mt_genes)  # Number of mitochondrial genes
```

> Does this look fine? There are 13 mitochondrial genes in the features/probes

# Calculate average mt percent per sample

```{r}
avg_mt_df <- data.frame(
  sample = names(seurat_list_feature_count_filtered),
  avg_percent_mt = sapply(seurat_list_feature_count_filtered, function(obj) 
  {
    mean(obj$percent.mt)
  })
)
```

```{r}
write.table(x = avg_mt_df, file = "01_mito_content_analysis/average_mit_exp.tsv", quote = F,
            sep = "\t", row.names = F)
```


# Plot number of cells retained at each cutoff 

```{r}
# Vector of cutoff thresholds 
plot_mt_retention <- function(seurat_list, obj_name)
{
  cutoffs <- seq(5, 40, by=5)
  obj = seurat_list[[obj_name]]
  # For each cutoff, count how many cells have percent.mt < cutoff
  retained_cells <- sapply(cutoffs, function(x) sum(obj$percent.mt < x))

  # Create a data frame for plotting
  df <- data.frame(cutoff = cutoffs, n_cells_retained = retained_cells)
  retained_20 <- df$n_cells_retained[df$cutoff == 20]
  total_cells <- ncol(obj)
# Plot
  p <- ggplot(df, aes(x = cutoff, y = n_cells_retained)) +
        geom_line(size = 1) +
        geom_point() +
        geom_vline(xintercept = 20, linetype = "dashed", color = "gray40") +
        geom_text(aes(x = 20, y = retained_20, label = paste0("n = ", retained_20)),
                vjust = -0.5, hjust = -0.1, color = "red", size = 5) +
        labs(title = paste0("Cells Retained at Different Mitochondrial Cutoffs ", obj_name, " (total cells = ", total_cells ,")" ),
           x = "Mitochondrial % Cutoff",
          y = "Number of Cells Retained") +
        theme_minimal()
  print(p)
}
```

```{r}
for (sample_name in names(seurat_list_feature_count_filtered))
{
  plot_mt_retention(seurat_list_feature_count_filtered, sample_name)
}
```


---
Data-driven approach to set the cutoff. Here the original data is used (unfiltered in any form)

Merge the seurat object

```{r}
merged_seurat <- merge(seurat_list[[1]],
                       seurat_list[-1],
                       add.cell.ids=names(seurat_list),
                       project="merged_unfiltered")
```

Number of cells per sample:
```{r}
table(sapply(strsplit(colnames(merged_seurat), "_"), `[`, 1)) 
```


# Remove empty cell barcodes
```{r}
merged_seurat <- subset(merged_seurat, subset = nFeature_RNA > 0 & nCount_RNA > 0)
```


Number of cells per sample after removing empty barcodes:
```{r}
table(sapply(strsplit(colnames(merged_seurat), "_"), `[`, 1)) 
```


percent mt

Calculate percent MT
```{r}
merged_seurat[["percent.mt"]] <- PercentageFeatureSet(merged_seurat, pattern = "^mt-")
```


Prelim quality: remove cells that are above 40% mito content
```{r}
merged_seurat <- subset(merged_seurat, subset = percent.mt < 40)
```

Number of cells per sample after removing > 40% mito content:
```{r}
table(sapply(strsplit(colnames(merged_seurat), "_"), `[`, 1)) 
```

```{r}
VlnPlot(merged_seurat, features = "percent.mt", group.by = "sample_id", raster=F) +
  geom_hline(yintercept = 20, linetype = "dashed", color = "red", size = 1) +
  geom_hline(yintercept = 25, linetype = "dashed", color = "blue", size = 1) +
  theme_minimal() +
  ggtitle("Mitochondrial % by Sample")

```


---

```{r}
merged_filtered_FC_seurat <- merge(seurat_list_feature_count_filtered[[1]],
                                   seurat_list_feature_count_filtered[-1],
                                   add.cell.ids=names(seurat_list_feature_count_filtered),
                                   project="merged")
```

```{r}
table(sapply(strsplit(colnames(merged_filtered_FC_seurat), "_"), `[`, 1)) 
```

```{r}
VlnPlot(merged_filtered_FC_seurat, features = "percent.mt", group.by = "sample_id") +
  geom_hline(yintercept = 20, linetype = "dashed", color = "red", size = 1) +
  geom_hline(yintercept = 25, linetype = "dashed", color = "blue", size = 1) +
  theme_minimal() +
  ggtitle("Mitochondrial % by Sample")

```

```{r}
VlnPlot(merged_filtered_FC_seurat, features = "nFeature_RNA", group.by = "sample_id") +
  geom_hline(yintercept = 1000, linetype = "dashed", color = "red", size = 1) +
  geom_hline(yintercept = 5000, linetype = "dashed", color = "blue", size = 1) +
  theme_minimal() +
  ggtitle("Feature counts by Sample")

```

```{r}
VlnPlot(merged_filtered_FC_seurat, features = "nCount_RNA", group.by = "sample_id") +
  geom_hline(yintercept = 1000, linetype = "dashed", color = "red", size = 1) +
  geom_hline(yintercept = 20000, linetype = "dashed", color = "blue", size = 1) +
  theme_minimal() +
  ggtitle("UMI counts by Sample")

```

```{r}
for (sample_name in names(seurat_list)) {
  seurat_obj <- seurat_list[[sample_name]]
  valid_cells <- !is.na(seurat_obj$percent.mt)
  VlnPlot(seurat_obj[, valid_cells], features = "percent.mt", alpha = 0, pt.size = 0)
}
```

```{r}
VlnPlot(seurat_list[["Ctrl72h"]], features = "percent.mt", alpha = 0, pt.size = 0)
```

```{r}
summary(seurat_list[["Ctrl72h"]]@meta.data$percent.mt)
```

```{r}
sample_Ctrl72h <- seurat_list[["Ctrl72h"]]
sample_Ctrl72h <- subset(sample_Ctrl72h, subset =  nFeature_RNA > 200)
summary(sample_Ctrl72h@meta.data$percent.mt)
summary(sample_Ctrl72h@meta.data$nCount_RNA)
```


```{r}
for (sample_name in names(seurat_list)) {
  cat(sample_name, " ",ncol(seurat_list[[sample_name]]), "\n")
}
```

```{r}
for (sample_name in names(seurat_list)) {
  seurat_list[[sample_name]] <- subset(seurat_list[[sample_name]], subset = nFeature_RNA > 200 & percent.mt < 10)
}
```


```{r}
for (sample_name in names(seurat_list)) {
  cat(sample_name, " ",ncol(seurat_list[[sample_name]]), "\n")
}
```

```{r}
for (sample_name in names(seurat_list)) {
  plot1 <- FeatureScatter(seurat_list[[sample_name]], feature1 = "nCount_RNA", feature2 = "percent.mt") + NoLegend()
  plot2 <- FeatureScatter(seurat_list[[sample_name]], feature1 = "nCount_RNA", feature2 = "nFeature_RNA") + NoLegend()
  print(plot1 + plot2 )
}
```


```{r}
# Convert to data frame for ggplot
rna_counts <- data.frame(nCount_RNA = seurat_list$`Ctrl72h`$nCount_RNA)

# Histogram
ggplot(rna_counts, aes(x = nCount_RNA)) +
  geom_histogram(bins = 100, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of RNA counts per cell",
       x = "RNA counts",
       y = "Number of cells")

# Density plot
ggplot(rna_counts, aes(x = nCount_RNA)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  theme_minimal() +
  labs(title = "Distribution of RNA counts per cell - Ctrl72h",
       x = "RNA counts",
       y = "Density")
```



