---
title: "02_XAL_mouse_sc"
output: html_document
date: "2025-05-29"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries

```{r echo=FALSE}
library(dplyr)
library(Seurat)
library(patchwork)
library(here)
library(ggplot2)
library(stringr)
```

## Analysis

### Keep only XAL

```{r}
h5_dir <- here("/Users/tharveshmoideliyakatali/Documents/XX_Steffan_UiO_RNA_seq/mouse_single_cell/individual_files")
h5_files <- list.files(h5_dir, pattern = "\\.h5$", full.names = TRUE)
h5_files <- h5_files[grep("XAL", h5_files)]

#Read metadata
meta_path <- here("../aggregation.csv")
metadata <- readr::read_csv(meta_path)
metadata <- metadata %>% filter(group=="XAL")
metadata$molecule_h5 <- NULL
```

## Reading count files

```{r}
seurat_list <- list()

for(file in h5_files)
{
  sample_name <- gsub(".h5$", "", basename(file))
  sample_name <- gsub("^[^_]+_", "", sample_name)
  counts <- Read10X_h5(file)
  seurat_obj <- CreateSeuratObject(counts = counts, project = sample_name)
  seurat_list[[sample_name]] <- seurat_obj
}
```
## Add metadata to the metadata

```{r}
for (sample_name in names(seurat_list)) {
  # Get corresponding metadata for this sample
  # Adjust this based on how your metadata is organized
  sample_metadata <- metadata %>% filter(sample_name == sample_id)
  
  # Add sample-level metadata to Seurat object
  for (col in colnames(sample_metadata)) {
    if (col != "sample") {
      seurat_list[[sample_name]][[col]] <- sample_metadata[[col]][1]
    }
  }
}
```

```{r}
for (sample_name in names(seurat_list)) {
  cat(sample_name, " ",ncol(seurat_list[[sample_name]]), "\n")
}
```

# Cell/barcode counts at the beginning

```{r}
cb_counts_start <- data.frame(
  sample = names(seurat_list),
  n_cb = sapply(seurat_list, function(obj) ncol(obj))
)
cb_counts_start
```

# Filter based on Jesse's cutoff

```{r}
min_features <- 200
max_features <- 7500
min_counts <- 400
max_counts <- 40000

seurat_list_feature_count_filtered <- lapply(seurat_list, function(obj) {
  obj <- subset(obj, subset = nFeature_RNA > min_features & 
                  nFeature_RNA <= max_features & 
                  nCount_RNA > min_counts &
                  nCount_RNA <= max_counts)
})
```

# Free up memory!
```{r}
rm(seurat_list)
gc()
```

# Cell/barcode counts after applying above cutoff

```{r}
cb_counts_filtered <- data.frame(
  sample = names(seurat_list_feature_count_filtered),
  n_cb = sapply(seurat_list_feature_count_filtered, function(obj) ncol(obj))
)
cb_counts_filtered
```

```{r}
cell_count_number <- left_join(cb_counts_start, cb_counts_filtered, by = 'sample')
colnames(cell_count_number) <- c("sample", "cb_wo_filter", "cb_filter_by_nfeatures_ncount")
write.table(cell_count_number, file = "04_XAL/cell_feature_umi_filtered.tsv", quote = FALSE,
            sep = "\t", row.names = FALSE)
```

# Calculate percent mt on the count filtered seurat object

```{r}
for (sample_name in names(seurat_list_feature_count_filtered)) {
  seurat_list_feature_count_filtered[[sample_name]][["percent.mt"]] <- PercentageFeatureSet(seurat_list_feature_count_filtered[[sample_name]], pattern = "^mt-")
}
```

# Calculate average mt percent per sample

```{r}
avg_mt_df <- data.frame(
  sample = names(seurat_list_feature_count_filtered),
  avg_percent_mt = sapply(seurat_list_feature_count_filtered, function(obj) 
  {
    mean(obj$percent.mt)
  })
)
```

```{r}
write.table(x = avg_mt_df, file = "04_XAL/average_mit_exp.tsv", quote = F,
            sep = "\t", row.names = F)
```

# Merge the seurat object

```{r}
merged_seurat_filtered_XAL <- merge(seurat_list_feature_count_filtered[[1]],
                                seurat_list_feature_count_filtered[-1],
                                add.cell.ids=names(seurat_list_feature_count_filtered),
                                project="gastruloid_merged_filtered")
```

# Plot mitochondrial gene percent

```{r}
VlnPlot(merged_seurat_filtered_XAL, features = "percent.mt", group.by = "sample_id", raster=F) +
  geom_hline(yintercept = 20, linetype = "dashed", color = "red", size = 1) +
  geom_hline(yintercept = 25, linetype = "dashed", color = "blue", size = 1) +
  theme_minimal() +
  ggtitle("Mitochondrial % by Sample")

```





```{r}
VlnPlot(merged_seurat_filtered_XAL, features = "nFeature_RNA", group.by = "sample_id") +
  geom_hline(yintercept = 151, linetype = "dashed", color = "red", size = 1) +
  geom_hline(yintercept = 7499, linetype = "dashed", color = "blue", size = 1) +
  theme_minimal() +
  ggtitle("Feature counts by Sample")

```

```{r}
VlnPlot(merged_seurat_filtered_XAL, features = "nCount_RNA", group.by = "sample_id") +
  geom_hline(yintercept = 400, linetype = "dashed", color = "red", size = 1) +
  geom_hline(yintercept = 40000, linetype = "dashed", color = "blue", size = 1) +
  theme_minimal() +
  ggtitle("UMI counts by Sample")

```

Remove cells with mt percent > 25%

```{r}
merged_seurat_filtered_XAL <- subset(merged_seurat_filtered_XAL, subset = percent.mt < 25)
```

Number of cells per sample after removing the high mito content:
```{r}
table(sapply(strsplit(colnames(merged_seurat_filtered_XAL), "_"), `[`, 1)) 
```

Write it to RDS
```{r}
saveRDS(merged_seurat_filtered_XAL, file = "04_XAL/04_XAL_sc_mouse_filtered_merged_seurat.rds")
```