---
title: "03_SC_processing_Jesse"
output: html_document
date: "2025-05-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load the libraries
```{r}
library(Seurat)
library(dplyr)
library(Matrix)
library(RColorBrewer)
library(pdist)
library(plyr)
library(reshape)
library(ggplot2)
library(reshape2)
library(openxlsx)
library(gplots)
library(corrplot)
library(viridis)
library(reshape2)
library(gridExtra)
library(cowplot)
library(reticulate)
library(umap)
library(monocle3)
library(SeuratWrappers)
library(patchwork)
library(viridis)
library(tidyverse)
library(smplot2)
#library(dittoSeq) 

```


```{r}
mouse_seurat_filtered <- readRDS('02_QC_mouse_single_cell/02_sc_mouse_filtered_merged_seurat.rds')
```


```{r}
VlnPlot(mouse_seurat_filtered, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```

```{r}
plot1 <- FeatureScatter(mouse_seurat_filtered, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(mouse_seurat_filtered, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2
```

# Combine all the identity

```{r}
# 1. Store current identity in a new metadata column
mouse_seurat_filtered$original_ident <- Idents(mouse_seurat_filtered)

# Unified identity
Idents(mouse_seurat_filtered) <- "gastruloid"
```


# Normalize and find the variable features

```{r}
mouse_seurat_filtered <- NormalizeData(mouse_seurat_filtered, normalization.method = "LogNormalize", scale.factor = 10000)

mouse_seurat_filtered <- FindVariableFeatures(mouse_seurat_filtered, selection.method = "vst", nfeatures = 2000)
```
save intermediate:
```{r}
saveRDS(mouse_seurat_filtered, file = "03_SC_preprocessing/03_sc_processing_j_normalised_intermediate.rds")
```

Read intermediate:
```{r}
mouse_seurat_filtered <- readRDS(file = "03_SC_preprocessing/03_sc_processing_j_normalised_intermediate.rds")
```



```{r}
# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(mouse_seurat_filtered), 10)
```


```{r}
# plot variable features with and without labels
plot1 <- VariableFeaturePlot(mouse_seurat_filtered)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot1 + plot2
```

```{r}
all.genes <- rownames(mouse_seurat_filtered)
mouse_seurat_filtered <- ScaleData(mouse_seurat_filtered, features = all.genes)
```

```{r}
mouse_seurat_filtered <- RunPCA(mouse_seurat_filtered, features = VariableFeatures(object = mouse_seurat_filtered))
```

```{r}
VizDimLoadings(mouse_seurat_filtered, dims = 1:2, reduction = "pca")
```

```{r}
DimHeatmap(mouse_seurat_filtered, dims = 1:15, cells = 500, balanced = TRUE)
```


```{r}
ElbowPlot(mouse_seurat_filtered, ndims = 30)
```

```{r}
mouse_seurat_filtered <- FindNeighbors(mouse_seurat_filtered, dims = 1:15)
mouse_seurat_filtered <- FindClusters(mouse_seurat_filtered)
```

```{r}
mouse_seurat_filtered <- RunUMAP(mouse_seurat_filtered, dims = 1:15)
```
```{r}
DimPlot(mouse_seurat_filtered, reduction = "pca")
```

```{r}
DimPlot(mouse_seurat_filtered, reduction = "umap", group.by = "original_ident")
```


```{r}
DimPlot(mouse_seurat_filtered, reduction = "umap", group.by = "time")
```

```{r}
DimPlot(mouse_seurat_filtered, reduction = "umap", group.by = "group")
```

```{r}
saveRDS(mouse_seurat_filtered, "03_SC_preprocessing/mouse_seurat_filtered_default_preprocessing.rds")
```

```{r}
mouse_seurat_filtered[["joined"]] <- JoinLayers(mouse_seurat_filtered[["RNA"]])
```



```{r}
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes
```


```{r}
mouse_seurat_filtered <- CellCycleScoring(mouse_seurat_filtered, s.features = s.genes, g2m.features = g2m.genes, set.ident = TRUE, assay="joined")
```

```{r}
DimPlot(mouse_seurat_filtered, reduction = "umap", group.by = "Phase")
```


Write metadata after calculating the cell cycle score
```{r}
write.table(mouse_seurat_filtered@meta.data, file = "03_SC_preprocessing/cell_cycle_scored_metadata.tsv",
            sep = "\t", quote = F)
```

Write the rdata
```{r}
saveRDS(mouse_seurat_filtered, file = "03_SC_preprocessing/mouse_seurat_filtered_default_preprocessing_cc.rds")
```

Remove joined layer
```{r}
mouse_seurat_filtered[["joined"]] <- NULL
```


Regressing the cell cycle score
```{r}
mouse_seurat_filtered <- ScaleData(mouse_seurat_filtered, vars.to.regress = c("S.Score", "G2M.Score"), features = rownames(mouse_seurat_filtered))
```

The step didn't work because of the memory issue. Trying to it per sample

```{r}
mouse_seurat_filtered_split <- SplitObject(mouse_seurat_filtered, split.by = "sample_id")
```

```{r}
mouse_seurat_filtered_split <- lapply(mouse_seurat_filtered_split, function(obj) {
  obj <- ScaleData(obj, vars.to.regress = c("S.Score", "G2M.Score"), features = rownames(obj))
  return(obj)
})
```

```{r}
mouse_seurat_filtered_cc <- merge(mouse_seurat_filtered_split[[1]], y = mouse_seurat_filtered_split[-1])
```

Removing the split seurat object

```{r}
rm(mouse_seurat_filtered_split)
gc()
```

Save the cell cycle regressed seurat object

```{r}
saveRDS(mouse_seurat_filtered_cc, 
        file = "03_SC_preprocessing/mouse_seurat_filtered_default_preprocessing_cc_regressed.rds")
```

```{r}
mouse_seurat_filtered_cc <- RunPCA(mouse_seurat_filtered_cc, features = VariableFeatures(mouse_seurat_filtered_cc), nfeatures.print = 10)
```


```{r}
DimPlot(mouse_seurat_filtered_cc, group.by="Phase")
```

THE ABOVE RESULT IS NOT DESIRED!!

Ran regression on a secured AWS network and copied to local

Remove what is currently loaded
```{r}
rm(mouse_seurat_filtered_cc)
gc()
```

Load what was processed on AWS

```{r}
mouse_seurat_filtered_cc_reg <- readRDS("03_SC_preprocessing/mouse_seurat_filtered_default_preprocessing_cc_regressed.rds")
```


```{r}
mouse_seurat_filtered_cc_reg <- RunPCA(mouse_seurat_filtered_cc_reg, features = VariableFeatures(mouse_seurat_filtered_cc_reg), nfeatures.print = 10)
```

FAILED BECAUSE OF MEMORY - Continuing on AWS.

Check phase info here to compare the plots

```{r}
mouse_seurat_filtered_cc <- readRDS("03_SC_preprocessing/mouse_seurat_filtered_default_preprocessing_cc_regressed_umap.rds")
```

```{r}
FeaturePlot(mouse_seurat_filtered_cc, features = "S.Score", reduction = "umap")
```

```{r}
FeaturePlot(mouse_seurat_filtered_cc, features = "G2M.Score", reduction = "umap")
```

```{r}
FeaturePlot(mouse_seurat_filtered_cc, features = "Utf1", reduction = "umap")
```

```{r}
DimPlot(mouse_seurat_filtered_cc, group.by = "Phase")
```

```{r}
dittoBarPlot(object = mouse_seurat_filtered_cc, group.by = "seurat_clusters",var  = "Phase", scale="percent")
```

```{r}
DimPlot(mouse_seurat_filtered_cc, group.by = "time")
```

```{r}
DimPlot(mouse_seurat_filtered_cc, group.by = "RNA_snn_res.0.5", label = T)
```

```{r}
DimPlot(mouse_seurat_filtered_cc, group.by = "original_ident")
```

```{r}
dittoBarPlot(object = mouse_seurat_filtered_cc, group.by = "seurat_clusters",var  = "original_ident", scale="percent")
```

```{r}
dittoBarPlot(object = mouse_seurat_filtered_cc, group.by = "seurat_clusters",var  = "time", scale="percent")
```

```{r}
DimPlot(mouse_seurat_filtered_cc, group.by = "RNA_snn_res.0.8", label = T)
```

```{r}
dittoBarPlot(object = mouse_seurat_filtered_cc, group.by = "RNA_snn_res.0.8",var  = "original_ident", scale="percent")
```

```{r}
dittoBarPlot(object = mouse_seurat_filtered_cc, group.by = "RNA_snn_res.0.8",var  = "time", scale="percent")
```

```{r}
dittoBarPlot(object = mouse_seurat_filtered_cc, group.by = "RNA_snn_res.0.8",var  = "group", scale="percent")
```

```{r}
cd_genes2 <- c("Zfp42","Dppa5a","Klf2","Sox2","Apoe","Fabp3","Cd63","Utf1","Pim2","Dnmt3b","Pou3f1","Otx2","Lefty1","Nodal","mt-Cytb","mt-Nd5","Epcam","Cdh1","Cdh2","T","Fgf5","Fgf8","Wnt3","Nkx1-2","Cdx2","Eomes","Mixl1","Gsc","Lhx1","Mesp1","Foxa2","Chrd","Cobl","Bicc1","Tbx6","Tcf15","Notch1","Dll1","Aldh1a2","Meox1","Foxc1","Prrx2","Osr1","Meis2","Hand1","Gata6","Lmo2","Kdr","Cdh5","Etv2","Sox17","Spink1","Amot","Cldn6","Ncam1","Sox1","Sox3","Sox11","Pax3","Crabp1","Tubb3")
cd_genes2 <- rev(cd_genes2)
DotPlot(mouse_seurat_filtered_cc, 
                    features = cd_genes2, group.by = "RNA_snn_res.0.5", 
                    cols = "Blues", scale.by = "size", 
                    scale = TRUE) + coord_flip() + scale_color_distiller(direction = 0)
```

```{r}
original_ident_order <- c("Ctrl48h", "RACL48h", "NACL48h", "XAL48h", 
                          "Ctrl72h", "RACL72h", "NACL72h", "XAL72h",
                          "NACL96h", "XAL96h", "NACL120h", "XAL120h")
mouse_seurat_filtered_cc$original_ident <- factor(mouse_seurat_filtered_cc$original_ident,
                                                  levels = original_ident_order)
DotPlot(mouse_seurat_filtered_cc, 
                    features = cd_genes2, group.by = "original_ident", 
                    cols = "Blues", scale.by = "size", 
                    scale = TRUE) + coord_flip() + scale_color_distiller(direction = 0)
```

```{r}
time_order <- c("48h", "72h", "96h", "120h")
mouse_seurat_filtered_cc$time <- factor(mouse_seurat_filtered_cc$time, levels = time_order)
DotPlot(mouse_seurat_filtered_cc, 
                    features = cd_genes2, group.by = "time", 
                    cols = "Blues", scale.by = "size", 
                    scale = TRUE) + coord_flip() + scale_color_distiller(direction = 0)
```

```{r}
FeaturePlot(mouse_seurat_filtered_cc, 
                 features = c("Zfp42","Dppa5a","Klf2","Sox2"), 
                 order = TRUE, cols = c("grey92", "dodgerblue4"), pt.size = 0.5)
```





